---
title: "Introductory Example - Drosophila Larval Olfaction"
author: "Gregory Jefferis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introductory Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=6
)
rgl::setupKnitr()
```

## Intro
This example is designed to introduce some of the basic features of
the catmaid package and the NeuroAnatomy Toolbox ([nat](https://natverse.github.io/nat/))
on which it builds.

The sample data is from an EM connectomics study of the Drosophila first instar larva
by [Albert Cardona and colleagues](https://elifesciences.org/articles/14859) as
submitted to https://virtualflybrain.org and visible at https://catmaid.virtualflybrain.org

## Connecting

First we need to specify which CATMAID server we will use. This
will be the default for the rest of the session.

```{r}
library(catmaid)
conn=catmaid_login(server="https://l1em.catmaid.virtualflybrain.org")
```

## Fetching neurons

Now lets fetch the peripheral olfactory sensory neurons, reading them into a 
`neuronlist()` object called `orns`.
```{r orns}
orns=read.neurons.catmaid("name:ORN (left|right)")
```

We can also add some useful metadata. 
```{r orn-meta}
orns[,'side']= factor(sub(".* ORN ", "", orns[,'name']))
orns[,'Or']= factor(sub(" ORN.*", "", orns[,'name']))
```

We calculate both the side of the brain and the [odorant receptor](https://en.wikipedia.org/wiki/Olfactory_receptor) expressed by the
olfactory sensory neurons. These two variables are stored as columns in a table (i.e. `data.frame()`) that is an attribute of the `orns` `neuronlist` object.

The *catmaid* package builds on [nat](https://natverse.github.io/nat), a general
purpose R neuroanatomy package. You may want to read the [nat documentation](https://natverse.github.io/nat/articles/neurons-intro.html) to
learn about how *nat* handles single neurons, collections of neurons (called `neuronlist`s)
and their associated metadata.

Now repeat for the the second order projection neuron (PN) partners.
Note use of search by annotation. See `catmaid_skids()` documentation for details.
```{r pns}
# searches for annotation "ORN PNs"
pns=read.neurons.catmaid("ORN PNs")
pns[,'Or']= factor(sub(" PN.*", "", pns[,'name']))
pns[,'side']= factor(sub(".*(left|right)", "\\1", pns[,'name']))
```

## Plotting neurons

Now let's plot both sets of neurons, colouring by odorant receptor.
```{r plot, rgl=T}
plot3d(orns, col=Or)
# note that we plot somata with a radius of 1500 nm
plot3d(pns, col=Or, soma=1500)
# set 3D viewer to frontal aspect
nview3d('frontal')
```

## Going further
### Finding synaptic partners
Now we can directly identify the downstream partner
neurons of the ORNs and plots them in 3D coloured by their synaptic strength.
It then carries out morphological clustering with [NBLAST](http://bit.ly/nblast)
and plots the partner neurons according to those clusters.

Let's find all the ORN downstream partners with at least 2 synapses
```{r partners}
orn_partners=catmaid_query_connected(orns[,'skid'], minimum_synapses = 2)
# keep the ones not already in our set of PNs
# there are lots!
non_pn_downstream_ids=setdiff(unique(orn_partners$outgoing$partner), pns[,'skid'])
# download and plot those neurons
non_pn_downstream=read.neurons.catmaid(non_pn_downstream_ids)
```

Note that although we exclude the downstream PNs 

```{r plot-downs, rgl=TRUE}
plot3d(orns, col=Or)
plot3d(non_pn_downstream, col='grey', soma=1000)
```

### Plot by connection strength
Plot, but colouring partners by number of synapses they receive from ORNs
First collect synapse numbers:
```{r}
library(dplyr)
totsyndf=orn_partners$outgoing %>% 
  group_by(partner) %>% 
  summarise(totsyn=sum(syn.count)) %>% 
  arrange(desc(totsyn))
hist(totsyndf$totsyn)
```
 
Now do the plot
```{r downs-strength, rgl=TRUE}
# matlab style palette
jet.colors <-
colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan",
"#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
# plot colouring by synapse number on a log scale 
# note that it is necessary to convert totsyndf$partner to a character
# vector to ensure that they are not treated as integer indices
plot3d(as.character(totsyndf$partner),  db=c(pns, non_pn_downstream), 
  col=jet.colors(10)[cut(totsyndf$totsyn, breaks = 2^(0:10))], soma=1000)
```


### NBLAST cluster

Now let's cluster these other connected neurons morphologically. This is
often an efficient way to group neurons that may have distinct functions.

```{r}
library(nat.nblast)
# convert to nblast-compatible format
# nb also convert from nm to um, resample to 1µm spacing and use k=5
# nearest neighbours of each point to define tangent vector
non_pn_downstream.dps=dotprops(non_pn_downstream/1e3, k=5, resample=1)
# now compute all x all NBLAST scores and cluster
non_pn_downstream.aba=nblast_allbyall(non_pn_downstream.dps)
non_pn_downstream.hc=nhclust(scoremat = non_pn_downstream.aba)
```

Plot result of clustering as dendrogram, labelled by abbreviated 
neuron name (rather than id).
```{r downstream-dendrogram}
plot(non_pn_downstream.hc, label=abbreviate(non_pn_downstream[,'name'], minlength = 8))
```

Plot in 3d cutting into 2 clusters essentially left right
```{r 2-clusters, rgl=T}
plot3d(non_pn_downstream.hc,db=non_pn_downstream, k=2, soma=1000)
```

Let's just try for one half of the brain. First we compute the side of
brain based on the neuron names.

```{r}
non_pn_downstream[,'side']='?'
non_pn_downstream[grepl("left", non_pn_downstream[,'name']),'side']='L'
non_pn_downstream[grepl("right", non_pn_downstream[,'name']),'side']='R'
```

Then we select and cluster just the neurons on the right hand side.

```{r 4-clusters}
rhsids=subset(non_pn_downstream, side == 'R', rval = 'name')
non_pn_downstream.hcr = nhclust(rhsids, scoremat = non_pn_downstream.aba)
# note that we abbreviate only the selected neuron names
# by indexing with the ids for the RHS neurons.
plot(
  non_pn_downstream.hcr,
  label = abbreviate(non_pn_downstream[rhsids, 'name'], minlength = 8))

```

Finally here is the plot with 4 clusters coloured.

```{r 4-clusters-plot, rgl=T}
plot3d(non_pn_downstream.hcr,db=non_pn_downstream, k=4, soma=1000)
```

